---
title: "R Notebook"
output: html_notebook
---

```{r}
selection <- list(
  regre_101 = list(yaml = "_bookdown_regression_101.yml", 
                   title = "Applications of Machine Learning. Regression 101"),
  regre_202 = list(yaml = "_bookdown_regression_202.yml",
                   title = "Applications of Machine Learning. Regression 202"),
  class_101 = list(yaml = "_bookdown_classification_101.yml",
                   title = "Applications of Machine Learning. Classification 101"),
  class_202 = list(yaml = "_bookdown_classification_202.yml",
                   title = "Applications of Machine Learning. Classification 202"),
  comp = list(yaml = "_bookdown_comparison.yml",
              title = "Comparison"),
  meta = list(yaml = "_bookdown_meta.yml",
              title = "Meta"),
  misc = list(yaml = "_bookdown_misc.yml",
              title = "Miscellaneous"),
  custom = list(yaml = "_bookdown_custom.yml",
                title = "Custom")
)

selection
```

```{r}
names(selection)
```

Select with the value that comes with `wicxh`

```{r}
which <- "regre_101"
selection[[which(names(selection) == "regre_101")]]
```

```{r}
selection[[which(names(selection) == "class_202")]]
```

```{r}
selection[[which(names(selection) == "comp")]]
```

When the name of the notebook is "comp"

```{r}
nbook <- "comp"
selection[[which(names(selection) == nbook)]]
```


## Find "title:" in `index.Rmd

1. Open the file `index.Rmd`
2. Apply regex to find the keyword `title:`
3. Extract that line
4. Get the sentence after the `:` as a string
5. Add the desired keyword to the sentence after a dot at the end of the sentence
6. Replace the line of text with `title:`
7. Save the file `index.Rmd`

```{r}
text <- readLines("index.Rmd")
```

### replace the current sentence as a pattern

```{r}
keyword <- "Regression"
sentence <- "Applications of Machine Learning"
new_text <- gsub(pattern = sentence, replace = paste0(sentence, ". ", keyword), 
                 x = text)
```

```{r}
writeLines(new_text, con = "index_new.Rmd")
```

### replace the whole title

```{r}
match_word <- "title:"
txt_line <- grep(pattern = match_word, x = text, value = TRUE)
txt_line
nLine <- grep(pattern = match_word, x = text)
nLine
```

```{r}
pat <- text[nLine]
pat
```

```{r}
orig_txt_line <- gsub(pattern = "title:", replacement = "", x = txt_line)
orig_txt_line <- trimws(gsub(pattern = '\"', replacement = "", x = orig_txt_line,  perl=TRUE))
new_txt_line
```


```{r}
new_sentence <- paste0("title: ", "\\'", "ML ", keyword, "\\'")
new_text <- gsub(pattern = pat, replace = new_sentence, 
                 x = text)
new_text[nLine]
```

```{r}
writeLines(new_text, con = "index_new.Rmd")
```

## Function

* Inputs
  - file index.Rmd
  - stop_at
  - replace 
* Stop if
  - 
* Actions  
  - replace string in file
  - save new text to file
* Return: list
  - success: TRUE or FALSE
  - line number when stopped
  - original string
  - replacement
  

```{r}
replace_in_index_rmd <- function(rmd_file, stop_word, replace, append = TRUE) {
    # stop if not Rmd file
    # stop if keyword not having colon
    space <- " "
    text <- readLines(rmd_file)
    nLine <- grep(pattern = stop_word, x = text)
    if (length(nLine) == 0) return(FALSE)
    pat <- text[nLine]
    txt_line <- grep(pattern = stop_word, x = text, value = TRUE) # get whole line
    orig_txt_line <- gsub(pattern = "title:", replacement = "", x = txt_line)
    orig_txt_line <- trimws(gsub(pattern = '\"', replacement = "", 
                                 x = orig_txt_line,  perl=TRUE)) # get rid of the stop_word
    if (append) {
      new_sentence <- paste0(stop_word, space, "\\'", orig_txt_line, space, 
                           replace, "\\'")  # append to current sentence
    } else {
      new_sentence <- paste0(stop_word, space, "\\'",  
                           replace, "\\'")  # make a new sentence
    }
    new_text <- gsub(pattern = pat, replace = new_sentence, x = text) # replace in line
    writeLines(new_text, con = rmd_file)
    # new_text[nLine]
}

replace_in_index_rmd("index.Rmd", "title:", "by NN")
replace_in_index_rmd("index.Rmd", "title:", "by NN", append = FALSE)
```


