# Diagnostic Plots

```{r setup, include=F, error=T, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE, 
                      comment =":>", 
                      error = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.align = 'center')
```

```{r echo=FALSE}
load(file.path(rprojroot::find_rstudio_root_file(), "workspace.RData"))
.libPaths(c(normalizePath(sbox_path), normalizePath(lib_path), .libPaths()))
```

I have tried to use fortify function in ggplot2 which can access different statistics related to linear model. The basic diagnostic plot which we often get using plot function in the fitted model using lm command.

The function `diagPlots` gives an list of six different plots which can be arranged in a grid using `grid` and `gridExtra` packages.

```{r}
library(ggplot2)

diagPlot<-function(model){
    p1<-ggplot(model, aes(.fitted, .resid))+geom_point()
    p1<-p1+stat_smooth(method="loess")+geom_hline(yintercept=0, col="red", linetype="dashed")
    p1<-p1+xlab("Fitted values")+ylab("Residuals")
    p1<-p1+ggtitle("Residual vs Fitted Plot")+theme_bw()
    
    p2<-ggplot(model, aes(qqnorm(.stdresid)[[1]], .stdresid))+geom_point(na.rm = TRUE)
    p2<-p2+geom_abline(aes(qqline(.stdresid)))+xlab("Theoretical Quantiles")+ylab("Standardized Residuals")
    p2<-p2+ggtitle("Normal Q-Q")+theme_bw()
    
    p3<-ggplot(model, aes(.fitted, sqrt(abs(.stdresid))))+geom_point(na.rm=TRUE)
    p3<-p3+stat_smooth(method="loess", na.rm = TRUE)+xlab("Fitted Value")
    p3<-p3+ylab(expression(sqrt("|Standardized residuals|")))
    p3<-p3+ggtitle("Scale-Location")+theme_bw()
    
    p4<-ggplot(model, aes(seq_along(.cooksd), .cooksd))+geom_bar(stat="identity", position="identity")
    p4<-p4+xlab("Obs. Number")+ylab("Cook's distance")
    p4<-p4+ggtitle("Cook's distance")+theme_bw()
    
    p5<-ggplot(model, aes(.hat, .stdresid))+geom_point(aes(size=.cooksd), na.rm=TRUE)
    p5<-p5+stat_smooth(method="loess", na.rm=TRUE)
    p5<-p5+xlab("Leverage")+ylab("Standardized Residuals")
    p5<-p5+ggtitle("Residual vs Leverage Plot")
    p5<-p5+scale_size_continuous("Cook's Distance", range=c(1,5))
    p5<-p5+theme_bw()+theme(legend.position="bottom")
    
    p6<-ggplot(model, aes(.hat, .cooksd))+geom_point(na.rm=TRUE)+stat_smooth(method="loess", na.rm=TRUE)
    p6<-p6+xlab("Leverage hii")+ylab("Cook's Distance")
    p6<-p6+ggtitle("Cook's dist vs Leverage hii/(1-hii)")
    p6<-p6+geom_abline(slope=seq(0,3,0.5), color="gray", linetype="dashed")
    p6<-p6+theme_bw()
    
    return(list(rvfPlot=p1, qqPlot=p2, sclLocPlot=p3, cdPlot=p4, rvlevPlot=p5, cvlPlot=p6))
}
```

Using the mtcars datasets, a linear model is fitted with mpg as response and cyl, disp, hp, drat and wt has predictor variable

```{r}
lm.model <- lm(mpg ~ cyl+disp+hp+drat+wt, data=mtcars)
diagPlts <- diagPlot(lm.model)
```

To display the plots in a grid, some packages mentioned above should be installed.

```{r}
lbry <- c("grid", "gridExtra")
lapply(lbry, require, character.only=TRUE, warn.conflicts = FALSE, quietly = TRUE)
```

Thus the plot obtained is,

```{r}
do.call(grid.arrange, c(diagPlts, main="Diagnostic Plots", ncol=3))
```

## Residual vs Fitted plot

```{r}
model <- lm.model
    p1<-ggplot(model, aes(.fitted, .resid)) + 
        geom_point() + 
        stat_smooth(method="loess") + 
        geom_hline(yintercept=0, col="red", linetype="dashed") +
        xlab("Fitted values") + 
        ylab("Residuals") +
        ggtitle("Residual vs Fitted Plot") + 
        theme_bw()
    p1
```

## Nomral QQ
```{r}
    p2 <- ggplot(model, aes(qqnorm(.stdresid)[[1]], .stdresid)) + 
    geom_point(na.rm = TRUE) +
    geom_abline(aes(qqline(.stdresid))) 
    # xlab("Theoretical Quantiles") + 
    # ylab("Standardized Residuals") + 
    # ggtitle("Normal Q-Q") + 
    # theme_bw()

p2
```

```{r}
    y <- quantile(model$resid[!is.na(model$resid)], c(0.25, 0.75))
    x <- qnorm(c(0.25, 0.75))
    slope <- diff(y)/diff(x)
    int <- y[1L] - slope * x[1L]
    p <- ggplot(model, aes(sample = .resid)) +
        stat_qq(alpha = 0.5) +
        geom_abline(slope = slope, intercept = int, color="blue") + 
        theme_bw()
    
    p
```


The standard Q-Q diagnostic for linear models plots quantiles of the standardized residuals vs. theoretical quantiles of N(0,1). Peter's ggQQ function plots the residuals. The snippet below amends that and adds a few cosmetic changes to make the plot more like what one would get from plot(lm(...)).

```{r}
# https://stackoverflow.com/a/19990107/5270873
ggQQ = function(lm) {
  # extract standardized residuals from the fit
  d <- data.frame(std.resid = rstandard(lm))
  # calculate 1Q/4Q line
  y <- quantile(d$std.resid[!is.na(d$std.resid)], c(0.25, 0.75))
  x <- qnorm(c(0.25, 0.75))
  slope <- diff(y)/diff(x)
  int <- y[1L] - slope * x[1L]

  p <- ggplot(data=d, aes(sample=std.resid)) +
    stat_qq(shape=1, size=3) +           # open circles
    labs(title="Normal Q-Q",             # plot title
         x="Theoretical Quantiles",      # x-axis label
         y="Standardized Residuals") +   # y-axis label
    geom_abline(slope = slope, intercept = int, linetype="dashed")  + # dashed reference line
      theme_bw()
  return(p)
}

ggQQ(model)
```


## Scale-location

```{r}
    p3 <- ggplot(model, aes(.fitted, sqrt(abs(.stdresid)))) + 
    geom_point(na.rm=TRUE) +
    stat_smooth(method="loess", na.rm = TRUE) + 
    xlab("Fitted Value") +
    ylab(expression(sqrt("|Standardized residuals|"))) + 
    ggtitle("Scale-Location") + 
    theme_bw()

    p3
```

## Cook's Distance
```{r}
    p4 <- ggplot(model, aes(seq_along(.cooksd), .cooksd)) + 
    geom_bar(stat="identity", position="identity") +
    xlab("Obs. Number") + 
    ylab("Cook's distance") +
    ggtitle("Cook's distance") + 
    theme_bw()

    p4

```

## Residual vs Leverage Plot

```{r}
    p5 <- ggplot(model, aes(.hat, .stdresid)) + 
    geom_point(aes(size=.cooksd), na.rm=TRUE) + 
    stat_smooth(method="loess", na.rm=TRUE) +
    xlab("Leverage") + 
    ylab("Standardized Residuals") +
    ggtitle("Residual vs Leverage Plot") +
    scale_size_continuous("Cook's Distance", range=c(1,5)) +
    theme_bw() + 
    theme(legend.position="bottom")

    p5
```

## Cook's dist vs Leverage hii/(1-hii)

```{r}
    p6 <- ggplot(model, aes(.hat, .cooksd)) + 
    geom_point(na.rm=TRUE) + 
    stat_smooth(method="loess", na.rm=TRUE) + 
    xlab("Leverage hii") + 
    ylab("Cook's Distance") +
    ggtitle("Cook's dist vs Leverage hii/(1-hii)") +
    geom_abline(slope=seq(0,3,0.5), color="gray", linetype="dashed") +
    theme_bw()

p6
```
