# Generalised additive models (GAMs): an introduction


## Introduction

Source: http://environmentalcomputing.net/intro-to-gams/

Many data in the environmental sciences do not fit simple linear models and are best described by "wiggly models", also known as Generalised Additive Models (GAMs).

Let’s start with a famous tweet by one Gavin Simpson, which amounts to:

1. GAMs are just GLMs
2. GAMs fit wiggly terms
3. use `+ s(x)` not `x` in your syntax
4. use method `= "REML"`
5. always look at `gam.check()`

This is basically all there is too it - an extension of generalised linear models (GLMs) with a smoothing function. Of course, there may be many sophisticated things going on when you fit a model with smooth terms, but you only need to understand the rationale and some basic theory. There are also lots of what would be apparently magic things happening when we try to understand what is under the hood of say lmer or glmer, but we use them all the time without reservation!

## Running the analysis
Before we consider a GAM, we need to load the package mgcv – the choice for running GAMs in R.

```{r}
library(mgcv)
library(ggplot2)
```


We’ll now look at a quick real example – we’ll just scratch the surface, and in a future tutorial we will look at it in more detail. We’re going to look at some CO2 data from Manua Loa (it’s used elsewhere in this series). We will fit a couple GAMs to the data to try and pick apart the intra- and inter-annual trends.

First load the data – you can download it here.

```{r}
CO2 <- read.csv(file.path(data_raw_dir, "mauna_loa_co2.csv"))
```

We want to look at inter-annual trend first, so let’s convert the date into a continuous time variable (take a subset for visualisation).

```{r}
CO2$time <- as.integer(as.Date(CO2$Date, format = "%d/%m/%Y"))
CO2_dat <- CO2
CO2 <- CO2[CO2$year %in% (2000:2010),]
```

OK, so let’s plot it and look at a smooth term for time.
$$y = \beta_0 + f_{\mathrm{trend}}(time) + \varepsilon, \quad \varepsilon \sim N(0, \sigma^2)$$


```{r}
ggplot(CO2_dat, aes(time, co2)) + geom_line()

```

We can fit a GAM for these data using:

```{r}
CO2_time <- gam(co2 ~ s(time), data = CO2, method = "REML")

```

which fits a model with a single smooth term for time. We can look at the predicted values for this:

```{r}
plot(CO2_time)

```

Note how the smooth term actually reduces to a ‘normal’ linear term here (with an edf of 1) – that’s the nice thing about penalised regression splines. But if we check the model, then we see something is amuck.

```{r}
par(mfrow = c(2,2))
gam.check(CO2_time)
```

The residual plots have a very odd looking rise-and-fall pattern – clearly there is some dependance structure (and we can probably guess it has something to do with intra-annual fluctuations). Let’s try again, and introduce something called a cyclical smoother.

$$ y = \beta_0 + f_{\mathrm{intrannual}}(month) + f_{\mathrm{trend}}(time) + \varepsilon, \quad \varepsilon \sim N(0, \sigma^2)$$

The cyclical smooth term, $f_{\mathrm{intrannual}}(month)$, is comprised of basis functions just the same as we have seen already, except that the end points of the spline are constrained to be equal – which makes sense when we’re modelling a variable that is cyclical (across months/years).

We’ll now see the bs = argument to choose the type of smoother, and the k = argument to choose the number of knots, because cubic regression splines have a set number of knots. We use 12 knots, because there are 12 months.


```{r}
CO2_season_time <- gam(co2 ~ s(month, bs = 'cc', k = 12) + s(time), 
                       data = CO2, 
                       method = "REML")

```

Let’s look at the fitted smooth terms:


```{r}
par(mfrow = c(1,2))
plot(CO2_season_time)
```

Looking at both smooth terms, we can see that the monthly smoother is picking up that monthly rise and fall of CO2 – looking at the relative magnitudes (i.e. monthly fluctuation vs. long-term trend), we can see how important it is to disintangle the components of the time series. Let’s see how the model diagnostics look now:

```{r}
par(mfrow = c(2,2))
gam.check(CO2_season_time)
```

Much better. Let’s look at how the seasonal component stacks up against the full long term trend.

```{r}
CO2_season_time <- gam(co2 ~ s(month, bs = 'cc', k = 12) + s(time), 
                       data = CO2_dat, 
                       method = "REML")
par(mfrow = c(1,2))
plot(CO2_season_time)
```
There’s more to the story – pephaps spatial autocorrelations of some kind? gam can make use of the spatial autocorrelation structures available in the nlme package, more on that next time. Hopefully for the meantime GAMs now don’t seem qutie so scary or magical, and you can start to make use of what is really an inrecibly flexible and powerful modelling framework.

## Communicating the results
You can essentially present model results from a GAM as if it were any other linear model, the main difference being that for the smooth terms, there is no single coefficient you can make inference from (i.e. negative, positive, effect size etc.). So you need to rely on either interpretting the parital effects of the smooth terms visually (e.g. from a call to plot(gam_model)) or make inference from the predicted values. You can of course include normal linear terms in the model (either continuous or categorical, and in an ANOVA type framework even) and make inference from them like you normally would. Indeed, GAMs are often useful for accounting for a non-linear phenomonon that is not directly of interest, but needs to be acocunted for when making inferece about other variables.

You can plot the partial effects by calling the plot function on a fitted gam model, and you can look at the parametric terms too, possibly using the termplot function too. You can use ggplot for simple models like we did earlier in this tutorial, but for more complex models, it’s good to know how to make the data using predict. We just use the existing time-series here, but you would generate your own data for the `newdata=` argument.

```{r}
CO2_pred <- data.frame(time = CO2_dat$time,
                       co2 = CO2_dat$co2,
                       predicted_values = predict(CO2_season_time, 
                                                  newdata = CO2_dat))
ggplot(CO2_pred, aes(x = time)) +
  geom_point(aes(y = co2), size = 1, alpha = 0.5) + 
    geom_line(aes(y = predicted_values), colour = "red")
```


