# jakub_glinka


## Multilayer perceptron
Multilayer perceptron (MLP) is the simplest feed-forward neural network. It mitigates the constraints of original perceptron that was able to learn only linearly separable patterns from the data. It achieves this by introducing at least one hidden layer in order to learn representation of the data that would enable linear separation.

Article: http://www.jakubglinka.com/2017-02-21-neural_networks_part1/
Code: https://github.com/jakubglinka/posts/blob/master/neural_networks_part1/neural_networks_part1.Rmd

## Requirements
* Install `gganimate` first
    devtools::install_github("dgrtwo/gganimate")
* Create folder `img`

## MXNet
MXNet is an open-source deep learning framework that allows you to define, train, and deploy deep neural networks on a wide array of devices, from cloud infrastructure to mobile devices and it allows to mix symbolic and imperative programming flavors. For example custom loss functions and accuracy measures.


## Network configuration
MXNet package expose so called symbolic API for R users. Its purpose is to create user friendly way of building neural networks abstracting out computational details to the MXNet specialized engine.

Most important symbols:

* `mx.symbol.Variable`: defines variables (input data, labels, …)
* `mx.symbol.FullyConnected`: affine transformation of input tensor
* `mx.symbol.Activation`: places activation function which is applied to all fields of input tensor
* `mx.symbol.Output`: defines final transformation of data and loss function.

Below is the example of code that configures perceptron with one hidden layer.

```{r}
 # required packages
  suppressPackageStartupMessages(library(ggplot2))
  library(tidyverse)
  library(lubridate)
  library(data.table)
  # suppressPackageStartupMessages(library(rstan))
  library(mxnet)
  library(mlbench)
  # require(gganimate)
  
# https://github.com/jakubglinka/posts/blob/master/files/mult_ggplot.R
source(file.path(r_code_dir, "mult_ggplot.R"))
```

## Universal Approximation Theorem

According to the theorem first proved by George Cybenko for sigmoid activation function: "feed-forward network with a single hidden layer containing a finite number of neurons (i.e., a multilayer perceptron), can approximate continuous functions on compact subsets of $\mathbb{R}^n$, under mild assumptions on the activation function."

Lets put mlp to the test then. For this purpose I will use sprials dataset from mlbench package.


```{r}
set.seed(2015)

  ############ sprials dataset ############

  s <- sample(x = c("train", "test"), 
              size = 1000, 
              prob = c(.8,.2),
              replace = TRUE)
  
  dta <- mlbench.spirals(n = 1000, cycles = 1.2, sd = .03)
  dta <- cbind(dta[["x"]], as.integer(dta[["classes"]]) - 1)
  colnames(dta) <- c("x","y","label")
  
  ######### train, validate, test ##########

  dta.train <- dta[s == "train",]
  dta.test <- dta[s == "test",]



  dta.plot <- data.frame(
    x = dta[, 1], 
    y = dta[, 2], 
    label = as.factor(dta[, 3]),
    sample = factor(s, levels = c("train", "test")))

  ggplot(aes(x = x, y = y), data = dta.plot) + 
    geom_point(aes(colour = label), size = 3) +
    geom_point(size = 3, shape = 1) + 
    theme_classic() + 
    ggtitle("Spirals dataset N = 1000")

```


```{r}

########### Network configuration ########
 
  # variables
  act <- mx.symbol.Variable("data")
  
  # affine transformation
  fc <- mx.symbol.FullyConnected(act, num.hidden = 10)
  
  # non-linear activation 
  act <- mx.symbol.Activation(data = fc, act_type = "relu")

  # affine transformation
  fc <- mx.symbol.FullyConnected(act, num.hidden = 2)
  
  # softmax output and cross-
  mlp <- mx.symbol.SoftmaxOutput(fc)
```


## Preparing data

```{r}
 set.seed(2015)

  ############ sprials dataset ############

  s <- sample(x = c("train", "test"), 
              size = 1000, 
              prob = c(.8,.2),
              replace = TRUE)
  
  dta <- mlbench.spirals(n = 1000, cycles = 1.2, sd = .03)
  dta <- cbind(dta[["x"]], as.integer(dta[["classes"]]) - 1)
  colnames(dta) <- c("x","y","label")
  
  ######### train, validate, test ##########

  dta.train <- dta[s == "train",]
  dta.test <- dta[s == "test",]
```


## Network training
Feed-forward networks are trained using iterative gradient descent type of algorithm. Additionally during single forward pass only subset of the data is used called batch. Process is repeated until all training examples are used. This is called an epoch. After every epoch MXNet returns training accuracy:
  
```{r}
  ############# basic training #############

  mx.set.seed(2014)
  model <- mx.model.FeedForward.create(
            symbol = mlp,
            X = dta.train[, c("x", "y")], 
            y = dta.train[, c("label")],
            num.round = 5,
            array.layout = "rowmajor",
            learning.rate = 1,
            eval.metric = mx.metric.accuracy)
```

## Custom call-back

In order to stop process of training when the progress in accuracy is below certain level of tolerance we need to add custom callback to the feed forward procedure. It is called after every epoch to check if algorithm progresses. If not it will terminate optimization procedure and return results.

```{r}
 ######## custom stopping criterion #######

  mx.callback.train.stop <- function(tol = 1e-3, 
                                     mean.n = 1e2, 
                                     period = 100, 
                                     min.iter = 100
                                     ) {
    function(iteration, nbatch, env, verbose = TRUE) {
      if (nbatch == 0 & !is.null(env$metric)) {
          continue <- TRUE
          acc.train <- env$metric$get(env$train.metric)$value
          if (is.null(env$acc.log)) {
            env$acc.log <- acc.train
          } else {
            if ((abs(acc.train - mean(tail(env$acc.log, mean.n))) < tol &
                abs(acc.train - max(env$acc.log)) < tol &
                iteration > min.iter) | 
                acc.train == 1) {
              cat("Training finished with final accuracy: ", 
                  round(acc.train * 100, 2), " %\n", sep = "")
              continue <- FALSE 
            }
            env$acc.log <- c(env$acc.log, acc.train)
          }
      }
      if (iteration %% period == 0) {
        cat("[", iteration,"]"," training accuracy: ", 
            round(acc.train * 100, 2), " %\n", sep = "") 
      }
      return(continue)
      }
   }

  ###### training with custom stopping #####

  mx.set.seed(2014)
  model <- mx.model.FeedForward.create(
          symbol = mlp,
          X = dta.train[, c("x", "y")], 
          y = dta.train[, c("label")],
          num.round = 2000,
          array.layout = "rowmajor",
          learning.rate = 1,
          epoch.end.callback = mx.callback.train.stop(),
          eval.metric = mx.metric.accuracy,
          verbose = FALSE
          )
```

## Results

### Learning curve

```{r}
grid <- 2:10
  res <- lapply(grid, function(xx) {
    
    act <- mx.symbol.Variable("data")
    fc <- mx.symbol.FullyConnected(act, num.hidden = xx)
    act <- mx.symbol.Activation(data = fc, act_type = "relu")
    fc <- mx.symbol.FullyConnected(act, num.hidden = 2)
    mlp <- mx.symbol.SoftmaxOutput(fc)
      
    mx.set.seed(2014)
    model <- mx.model.FeedForward.create(
      symbol = mlp,
      X = dta.train[, c("x", "y")], 
      y = dta.train[, c("label")],
      num.round = 2000,
      array.layout = "rowmajor",
      learning.rate = 1,
      epoch.end.callback = mx.callback.train.stop(),
      eval.metric = mx.metric.accuracy,
      verbose = FALSE
    )
    
    pred.test <- apply(
      predict(
        model, 
        X = dta.test[, c("x", "y")], 
        array.layout = "rowmajor"
        ), 2, which.max)
    
    test.acc <- e1071::classAgreement(table(pred.test, dta.test[, "label"]))[[1]]
    
    pred.train <- apply(
      predict(
        model, 
        X = dta.train[, c("x", "y")], 
        array.layout = "rowmajor"
        ), 2, which.max)
    
    train.acc <- e1071::classAgreement(table(pred.train, dta.train[, "label"]))[[1]]
    
    logp <- t(log(predict(
        model, 
        X = dta.test[, c("x", "y")], 
        array.layout = "rowmajor"
        )))
    loss <- - mean(logp[, 1] * (dta.test[, "label"] == 0) + 
                   logp[, 2] * (dta.test[, "label"] == 1))
    
    return(list(
      train = train.acc, 
      test = test.acc,
      loss = loss,
      model = model))
    
  })
  
```

```{r}
library(ggplot2)
library(dplyr)
library(reshape2)

tab <- res %>% 
    lapply(function(xx) data.frame(xx[c("train", "test")])) %>%
    do.call(rbind, .)
  tab$hidden.nodes <- grid
  
  tab <- tab %>% melt(id.vars = c("hidden.nodes"), 
                      measure.vars = c("train", "test"))
  names(tab) <- c("hidden.nodes", "sample", "accuracy")
  
  ggplot(aes(x = hidden.nodes), data = tab) + 
    geom_line(aes(y = accuracy, colour = sample), linetype = 2) + 
    geom_point(aes(y = accuracy, colour = sample, shape = sample, size = 3)) + 
    theme_classic() + 
    scale_y_continuous(labels = function(xx) paste0(round(xx * 100, 1), "%")) +
    ggtitle("Accuracy vs number of hidden nodes") + 
    guides(size = FALSE) +
    scale_size(range=c(0.8, 3))
  
```


### Evolution of decision boundary

```{r}
 grid <- expand.grid(x = seq(-1.1, 1.1, by = 0.05),
                      y = seq(-1.1, 1.1, by = 0.05))

  dtas <- lapply(1:length(res), function(xx) {
    
    prob <- predict(res[[xx]]$model, 
                    X = as.matrix(grid), 
                    array.layout = "rowmajor")
    prob <- t(prob)
    pred_label <- apply(prob, 1, which.max) - 1 
    
    require(dplyr)
    dta.plot <- bind_rows(data.frame(grid,
                                      prob = prob[, 1],
                                      cls="0",
                                      stringsAsFactors = FALSE),
                           data.frame(grid,
                                      prob = prob[, 2],
                                      cls="1",
                                      stringsAsFactors = FALSE)
    )
    dta.plot$frame <- xx - 1
    dta.plot
  })
  
  dta.plot <- bind_rows(dtas)
  
  p <- ggplot(aes(frame = frame), data = dta.plot) +
    geom_point(aes(x=x, y=y, col=cls, size = prob, frame = frame),
               alpha = .7,
               data = dta.plot[dta.plot$prob >= .5,]) + scale_size(range=c(0.8, 2)) + 
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 0,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 1,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 2,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 3,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 4,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 5,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 6,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 7,]) +
    geom_contour(aes(x=x, y=y, z= as.numeric(prob > .5), group=cls, color=cls),
                 bins=2, data = dta.plot[dta.plot$frame == 8,]) +
    theme_classic() +
    geom_point(aes(x=x, y=y, col=cls),
               size=3,
               data=data.frame(x=dta.train[,1], 
                               y=dta.train[,2],
                               frame = NA,
                               cls = as.character(dta.train[, 3]))) +
    geom_point(aes(x=x, y=y),
               size=3, shape=1,
               data=data.frame(x=dta.train[,1], 
                               y=dta.train[,2],
                               frame = NA,
                               cls = as.character(dta.train[, 3]))) + 
    ggtitle("Multilayer Perceptron decision boundary")
  
   
  
  # animation::ani.options(interval=2)
  # gganimate(p, filename = "{{ site.url }}/img/mlp_evolution.gif", title_frame = FALSE)
  # gganimate(p, filename = "img/mlp_evolution.gif", title_frame = FALSE)
```

```{r}
library(animation)
library(gganimate)
```



